Event History Analyses for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows

### 2024-10-24 manuscript/manuscript.Rmd
update ms. (add refs and format.args)

  <!-- https://tinyurl.com/ybremelq -->
keywords          : "response times, event history analysis, Bayesian multi-level regression models, experimental psychology, cognitive psychology"
wordcount         : "X"
bibliography      : ["extrareferences.bib", "r-references.bib"]
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "man"
appendix          : SupplementaryMaterial.Rmd
output            : papaja::apa6_pdf

@@

From a historical perspective, it is worth noting that the development of analytical tools that can estimate or predict whether and when events will occur is not a new endeavour.
Indeed, hundreds of years ago, analytical methods were developed to predict time to death (REFs).
The same logic has been applied to psychological time-to-event data, as previously demonstrated (Panis et al., 2020).
Here, in the paper, we hope to show the value of event history analysis for knowledge and theory building in cognitive psychology and related areas of research, such as cognitive neuroscience, as well as provide practical tutorials that provide step-by-step code and instructions in the hope that we can enable others to use event history analysis in a more routine, efficient and effective manner.

@@

In summary, even though event history analysis is a widely used statistical tool and there already exist many excellent reviews (REFs) and tutorials [@allisonSurvivalAnalysisUsing2010] on its general use-cases, we are not aware of any tutorials that are aimed at psychological time-to-event data, and which provide worked examples of the key data processing and multi-level regression modelling steps. 
Thereofore, our ultimate goal is twofold: first, we want to convince readers of the many benefits of using event history analysis when dealing with time-to-event data with a focus on psychological time-to-event data, and second, we want to provide a set of practical tutorials, which provide step-by-step instructions on how you actually perform a discrete-time event history analysis on time-to-event data such as RT data, as well as a complementary discrete-time SAT analysis on timed accuracy data.

@@

Discrete-time EHA focuses on the discrete-time hazard function of event occurrence and the discrete-time survivor function (Figure 2). The equations that define both of these functions are reported in the supplementary material (equations 1 and 2 in part A). The discrete-time hazard function gives you for each bin the probability that the event occurs (sometime) in bin t, given that the event does not occur in previous bins. In other words, it reflects the instantaneous likelihood that the event occurs in the current bin, given that it has not yet occurred in the past, i.e., in one of the prior bins. This conditionality in the definition of hazard is what makes the hazard function so diagnostic for studying event occurrence, as an event can physically not occur when it has already occurred before. In contrast, the discrete-time survivor function cumulates the bin-by-bin risks of event *non*occurrence to obtain the probability that the event occurs after bin t. In other words, the survivor function reflects the likelihood that the event occurs in the future, i.e., in one of the subsequent timebins.

[[this paragraph is not so clear yet... Perhaps add error bars to the functions in Figure 2?]]
The survivor function can help to qualify or provide context to the interpretation of the hazard function. For example, it can give a sense of how many trials contribute to each part of the hazard distribution. If a participant completes 100 trials in an experiment, and the survivor function reaches a probability of 0.03 at the end of timebin (400,500], then only 3% of trials remain beyond this point, which in this case would amount to 3 trials. Therefore, the error bars in later parts of the hazard function would be wider and less precise compared to earlier parts.

@@

In the context of experimental psychology, it is common for participants to be presented with either a 1-button detection task or a 2-button discrimination task, i.e., a task that has a right and a wrong answer. For example, a task may involve choosing between two response options with only one of them being correct. For such two-choice RT data, the discrete-time EHA can be extended with a discrete-time SAT analysis. Specifically, the hazard function of event occurrence can be extended with the discrete-time conditional accuracy function (see equation 5 in part A of the supplementary material), which gives you the probability that a response is correct given that it is emitted in time bin t [@kantowitzInterpretationReactionTime2021; @wickelgrenSpeedaccuracyTradeoffInformation1977; @allisonSurvivalAnalysisUsing2010]. We refer to this extended analysis for choice RT data as EHA/SAT.

@@

Note that the censoring time should be a multiple of the bin width (both in ms). The censoring time should be a time point after which no informative responses are expected anymore. In experiments that implement a response deadline in each trial the censoring time can equal that deadline time point. Trials with a RT larger than the censoring time, or trials in which no response is emitted during the data collection period, are treated as right-censored observations in EHA. In other words, these trials are not discarded, because they contain the information that the event did not occur before the censoring time. Removing such trials before calculating the mean event time can introduce a sampling bias (REFs).

@@

The next step is to set up the life table -- custom function setup_lt() --, calculate the conditional accuracies -- custom function calc_ca() --, and then plot the descriptive statistics using our custom function plot_eha(). When creating the plots, some warning messages will likely be generated, like these:

@@

Based on previous work [@panisHowCanWe2020; @panisNeuropsychologicalEvidenceTemporal2017; @panisStudyingDynamicsVisual2020; @panisTimecourseContingenciesPerceptual2009; @panisWhenDoesInhibition2022], we relax all three assumptions in model 4. We thus add a squared term for the continuous predictor centered trial number -- I(trial_c^2) -- and include interaction terms.

@@

```{r int-table}
apa_table(
  int_tab,
  caption = "(ref:int-table-caption)",
  note = "(ref:int-table-note-caption)",
  placement = "H",
  font_size = "scriptsize",
  longtable = TRUE,
  repeat_header = TRUE
)
```

@@

In the middle of the experiment (trial 1000), congruent and incongruent primes have positive effects in bins (200,240] and (240,280], while incongruent primes again have negative effects in bins (320,360], (360,400], and (400,440]. Due to practice, the primes generate a higher hazard of response occurrence for 80 ms (compared to 40 ms at the beginning of the experiment).

@@

```{r int-ca-table}
apa_table(
  int_ca_tab,
  caption = "(ref:int-ca-table-caption)",
  note = "(ref:int-ca-table-note-caption)",
  placement = "H",
  font_size = "scriptsize",
  longtable = TRUE,
  repeat_header = TRUE
)
```

@@

In this fifth tutorial we illustrate how to fit a multilevel hazard regression model in the frequentist framework, for the data set used in Tutorial 1a. The general process is similar to that in Tutorial 2a, except that there are no priors to set. For illustration purposes, we only fitted the effects from model M3 (see Tutorial 2a) using the function glmer() from the R package lme4. Alternatively, one could also use the function glmmPQL() from the R package MASS (REF). The resulting hazard model is called M3_f.

@@

(ref:descr-plot10-caption) Parameter estimates for model M3 from brm() and model M3_f from glmer().

@@

In this sixth tutorial we illustrate how to fit a multilevel regression model to the timed accuracy data in the frequentist framework, for the data set used in Tutorial 1a. For illustration purposes, we only fitted the effects from model M3 (see Tutorial 2a) using the function glmer() from the R package lme4. Alternatively, one could also use the function glmmPQL() from the R package MASS. Again, the resulting conditional accuracy model M3_ca_f did not converge and resulted in a singular fit.

@@

Models for discrete-time h(t) and ca(t) can quickly become very complex when adding more than 1 time scale, due to the many possible higher-order interactions. 
For example, model M4 contains two time scales as covariates: the passage of time on the across-bin or within-trial time scale (variable period_9), and the passage of time on the across-trial or within-experiment time scale (variable trial_c). However, when trials are presented in blocks, and blocks of trials within sessions, and when the experiment comprises three sessions, then four time scales can be defined (across-bin or within-trial, across-trial or within-block, across-block or within-session, and across-session or within-experiment).
From a theoretical perspective, adding more than 1 time scale is important to capture plasticity (e.g., proactive control) and other learning effects that play out on such longer time scales (across-trials, across-blocks, across-sessions), and that are probably present in each experiment in general.
From a practical perspective, therefore, it might be interesting for interpretational purposes to limit the number of experimental variables, because adding time scales quickly increases model complexity.

@@

Another approach: fit models to individual subjects and describe prevalence...REF

@@

The hazard models in this tutorial assume that there is one event of interest. For RT data, this event constitutes a single transition between an "idle" state and a "responded" state. However, in certain situations, more than one event of interest might exist. For example, in a medical or health-related context, an individual might transition back and forth between a "healthy" state and a "depressed" state, before a final "death" state. When you have data on the timing of these transitions, one can apply multi-state models, which generalize event history analysis to transitions between three or more states [@SteeleF2004].
Also, the predictor variables in this tutorial are time-invariant, i.e., their value did not change over the course of a trial. Thus, another extension is to include time-varying predictors, i.e., predictors whose value can change across the time bins within a trial [@allisonSurvivalAnalysisUsing2010]. For example, when gaze position is tracked during a visual search trial, the gaze-target distance will vary during a trial when the eyes move around before a manual response is given; shorter gaze-target distances should be associated with a higher hazard of response occurrence. Note that the effect of a time-varying predictor (e.g., an occipital EEG signal) can itself vary over time.

>>>>

  <!-- https://tinyurl.com/ybremelq -->
keywords          : "response times, event history analysis, Bayesian multi-level regression models, experimental psychology, cognitive psychology"
wordcount         : "X"
bibliography      : ["extrareferences2.bib", "r-references.bib"]
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "man"
appendix          : SupplementaryMaterial.Rmd
output            : papaja::apa6_pdf

@@

From a historical perspective, it is worth noting that the development of analytical tools that can estimate or predict whether and when events will occur is not a new endeavour.
Indeed, hundreds of years ago, analytical methods were developed to predict time to death [@williammatthewmakehamLawMortalityConstruction1860].
The same logic has been applied to psychological time-to-event data, as previously demonstrated (Panis et al., 2020).
Here, in the paper, we hope to show the value of event history analysis for knowledge and theory building in cognitive psychology and related areas of research, such as cognitive neuroscience, as well as provide practical tutorials that provide step-by-step code and instructions in the hope that we can enable others to use event history analysis in a more routine, efficient and effective manner.

@@

In summary, even though event history analysis is a widely used statistical tool and there already exist many excellent reviews [e.g., @blossfeldTechniquesEventHistory2002; @hosmerAppliedSurvivalAnalysis2011; @teachmanAnalyzingSocialProcesses1983; @box-steffensmeierEventHistoryModeling2004] and tutorials [e.g., @allisonSurvivalAnalysisUsing2010; @landesIntroductionEventHistory2020] on its general use-cases, we are not aware of any tutorials that are aimed at psychological time-to-event data, and which provide worked examples of the key data processing and multi-level regression modelling steps. 
Thereofore, our ultimate goal is twofold: first, we want to convince readers of the many benefits of using event history analysis when dealing with time-to-event data with a focus on psychological time-to-event data, and second, we want to provide a set of practical tutorials, which provide step-by-step instructions on how you actually perform a discrete-time event history analysis on time-to-event data such as RT data, as well as a complementary discrete-time SAT analysis on timed accuracy data.

@@

Discrete-time EHA focuses on the discrete-time hazard function of event occurrence and the discrete-time survivor function (Figure 2). The equations that define both of these functions are reported in the supplementary material (equations 1 and 2 in part A). The discrete-time hazard function gives you for each bin the probability that the event occurs (sometime) in bin t, given that the event does not occur in previous bins. In other words, it reflects the instantaneous likelihood that the event occurs in the current bin, given that it has not yet occurred in the past, i.e., in one of the prior bins. This conditionality in the definition of hazard is what makes the hazard function so diagnostic for studying event occurrence, as an event can physically not occur when it has already occurred before. In contrast, the discrete-time survivor function cumulates the bin-by-bin risks of event *non*occurrence to obtain the probability that the event occurs after bin t. In other words, the survivor function reflects the likelihood that the event occurs in the future, i.e., in one of the subsequent timebins. As explained in part A of the supplementary material, the survivor function provides a context for interpreting the hazard function.

<!-- The survivor function can help to qualify or provide context to the interpretation of the hazard function. For example, it can give a sense of how many trials contribute to each part of the hazard distribution. If a participant completes 100 trials in an experiment, and the survivor function reaches a probability of 0.03 at the end of timebin (400,500], then only 3% of trials remain beyond this point, which in this case would amount to 3 trials. Therefore, the error bars in later parts of the hazard function would be wider and less precise compared to earlier parts.-->

@@

In the context of experimental psychology, it is common for participants to be presented with either a 1-button detection task or a 2-button discrimination task, i.e., a task that has a right and a wrong answer. For example, a task may involve choosing between two response options with only one of them being correct. For such two-choice RT data, the discrete-time EHA can be extended with a discrete-time SAT analysis. Specifically, the hazard function of event occurrence can be extended with the discrete-time conditional accuracy function (see equation 5 in part A of the supplementary material), which gives you the probability that a response is correct given that it is emitted in time bin t [@kantowitzInterpretationReactionTime2021; @wickelgrenSpeedaccuracyTradeoffInformation1977; @allisonSurvivalAnalysisUsing2010]. We refer to this extended analysis for choice RT data as EHA/SAT. As explained in part A of the supplementary material, the probability mass function provides a context for interpreting the conditional accuracy function.

@@

Note that the censoring time should be a multiple of the bin width (both in ms). The censoring time should be a time point after which no informative responses are expected anymore. In experiments that implement a response deadline in each trial the censoring time can equal that deadline time point. Trials with a RT larger than the censoring time, or trials in which no response is emitted during the data collection period, are treated as right-censored observations in EHA. In other words, these trials are not discarded, because they contain the information that the event did not occur before the censoring time. Removing such trials before calculating the mean event time will result in underestimation of the true mean.

@@

The next step is to set up the life table using our custom function setup_lt(), calculate the conditional accuracies using our custom function calc_ca(), add the ca(t) estimates to the life table using our custom function join_lt_ca(), and then plot the descriptive statistics using our custom function plot_eha(). When creating the plots, some warning messages will likely be generated, like these:

@@

Based on previous work [@panisHowCanWe2020; @panisNeuropsychologicalEvidenceTemporal2017; @panisStudyingDynamicsVisual2020; @panisTimecourseContingenciesPerceptual2009; @panisWhenDoesInhibition2022] we expect nonlinear effects of trial number. We thus relax all three assumptions in model 4. We add a squared term for the continuous predictor centered trial number -- I(trial_c^2) -- and include interaction terms.

@@

```{r int-table}
apa_table(
  int_tab,
  caption = "(ref:int-table-caption)",
  note = "(ref:int-table-note-caption)",
  placement = "H",
  font_size = "scriptsize",
  longtable = TRUE,
  repeat_header = TRUE,
  format.args = list(
       digits = c(0, 2,2,2,2,2,5),
        decimal.mark = ".", big.mark = "")
)
```

@@

In the middle of the experiment (trial 1000), congruent and incongruent primes have positive effects in bins (200,240] and (240,280], while incongruent primes again have negative effects in bins (320,360], (360,400], and (400,440]. Due to practicing stimulus-response associations, the primes generate a higher hazard of response occurrence for 80 ms (compared to 40 ms at the beginning of the experiment).

@@

```{r int-ca-table}
apa_table(
  int_ca_tab,
  caption = "(ref:int-ca-table-caption)",
  note = "(ref:int-ca-table-note-caption)",
  placement = "H",
  font_size = "scriptsize",
  longtable = TRUE,
  repeat_header = TRUE,
  format.args = list(
       digits = c(0, 2,2,2,2,2,13),
        decimal.mark = ".", big.mark = "")
)
```

@@

In this fifth tutorial we illustrate how to fit a multilevel hazard regression model in the frequentist framework, for the data set used in Tutorial 1a. The general process is similar to that in Tutorial 2a, except that there are no priors to set. For illustration purposes, we only fitted the effects from model M3 (see Tutorial 2a) using the function glmer() from the R package lme4. Alternatively, one could also use the function glmmPQL() from the R package MASS [@ripleyMASSSupportFunctions2024]. The resulting hazard model is called M3_f.

@@

(ref:descr-plot10-caption) Parameter estimates for model M3 from brm() -- means and 95% quantile intervals -- and model M3_f from glmer() -- maximum likelihood estimates and 95% confidence intervals.

@@

In this sixth tutorial we illustrate how to fit a multilevel regression model to the timed accuracy data in the frequentist framework, for the data set used in Tutorial 1a. For illustration purposes, we only fitted the effects from model M3 (see Tutorial 2a) using the function glmer() from the R package lme4. Alternatively, one could also use the function glmmPQL() from the R package MASS [@ripleyMASSSupportFunctions2024]. Again, the resulting conditional accuracy model M3_ca_f did not converge and resulted in a singular fit.

@@

Models for discrete-time h(t) and ca(t) can quickly become very complex when adding more than 1 time scale, due to the many possible higher-order interactions. 
For example, model M4 contains two time scales as covariates: the passage of time on the across-bin or within-trial time scale (variable period_9), and the passage of time on the across-trial or within-experiment time scale (variable trial_c). However, when trials are presented in blocks, and blocks of trials within sessions, and when the experiment comprises three sessions, then four time scales can be defined (across-bin or within-trial, across-trial or within-block, across-block or within-session, and across-session or within-experiment).
From a theoretical perspective, adding more than 1 time scale -- and their interactions -- is important to capture plasticity and other learning effects (e.g., proactive control) that play out on such longer time scales (across-trials, across-blocks, across-sessions), and that are probably present in each experiment in general.
From a practical perspective, therefore, it might be interesting for interpretational purposes to limit the number of experimental predictor variables, because adding time scales quickly increases model complexity.

@@

Another approach to deal with individual differences is Bayesian prevalence [@inceBayesianInferencePopulation2021]. This method looks at effects within each individual in the study and asks how likely it would be to see the same result if the experiment was repeated with a new person chosen from the wider population at random. This approach allows one to quantify how typical or uncommon an observed effect is in the population, and the uncertainty around this estimate.

@@

The hazard models in this tutorial assume that there is one event of interest. For RT data, this event constitutes a single transition between an "idle" state and a "responded" state. However, in certain situations, more than one event of interest might exist. For example, in a medical or health-related context, an individual might transition back and forth between a "healthy" state and a "depressed" state, before a final "death" state. When you have data on the timing of these transitions, one can apply multi-state models, which generalize event history analysis to transitions between three or more states [@steeleGeneralMultilevelMultistate2004].
Also, the predictor variables in this tutorial are time-invariant, i.e., their value did not change over the course of a trial. Thus, another extension is to include time-varying predictors, i.e., predictors whose value can change across the time bins within a trial [@allisonSurvivalAnalysisUsing2010]. For example, when gaze position is tracked during a visual search trial, the gaze-target distance will vary during a trial when the eyes move around before a manual response is given; shorter gaze-target distances should be associated with a higher hazard of response occurrence. Note that the effect of a time-varying predictor (e.g., an occipital EEG signal) can itself vary over time.
